---
title: "simplified"
format: html
---

Simple model build to familiarize myself with concepts. I was getting lost in the process doing the SP500. Will scale up after I get the basics down.

```{r}
#packages
library(tidyquant)
library(tidyverse)
library(dplyr)
```

```{r}
# download market returns for portfolio --> lets just use 3-4 stocks to keep it simple
# [ (n^2 - n) / 2 ] + n days, where N = number stocks in portfolio
# GOOG, BG, RTX --> different sectors so have even less correlation (in theory)


goog <- tq_get("GOOG")
bg <- tq_get("BG")
rtx <- tq_get("RTX")

goog_1 <- goog |> mutate(daily_ret = ((adjusted - lag(adjusted))/lag(adjusted))*100)
bg_1 <- bg |> mutate(daily_ret = ((adjusted - lag(adjusted))/lag(adjusted))*100)
rtx_1 <- rtx |> mutate(daily_ret = ((adjusted - lag(adjusted))/lag(adjusted))*100)

head(goog_1)
head(bg_1)
head(rtx_1)
```


```{r}
returns_wide <- goog_1 |> select(date, GOOG = daily_ret) |>
  inner_join(bg_1 |> select(date, BG = daily_ret), by = "date") |>
  inner_join(rtx_1 |> select(date, RTX = daily_ret), by = "date") |>
  drop_na()
```

```{r}
#standarise the data. call it Z.
Z <- returns_wide |> select(-date) |> scale()
```

```{r}
# create correlation matrix from Z call it cor_Z. can skip cov step.
cor_Z <- cor(Z)
```

```{r}
# calculate eigenvalues and eigenvectors from corr_z
eigen <- eigen(cor_Z)

eigenvalues  <- eigen$values
eigenvectors <- eigen$vectors

var_explained     <- eigenvalues / sum(eigenvalues)
cum_var_explained <- cumsum(var_explained)
```

```{r}
#update: line graph not to be used. scree plot is for eigenvalues 
#https://www.r-bloggers.com/2023/10/creating-a-scree-plot-in-base-r/
#update: use tibble() not data.frame() for cleaner/better output.
#https://www.listendata.com/2024/05/tibble-vs-dataframe-with-examples.html

scree_df <- tibble(
  PC = paste0("PC", 1:length(eigenvalues)),
  num_pc = 1:length(eigenvalues),
  VarExplained = var_explained,
  CumVar = cum_var_explained
)

ggplot(scree_df, aes(x = num_pc, y = VarExplained)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "red", size = 3) +
  scale_x_continuous(breaks = 1:3) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Percent Variance Explained",
    x = "Eigenvector # (Principal Component)",
    y = "% Variance Explained"
  ) +
  theme_classic()
```


```{r}
# choose k eigenvectors based on previous outputs.
ggplot(scree_df, aes(x = num_pc, y = VarExplained)) +
  geom_line(color = "grey", size = 1) +
  geom_point(color = "grey42", size = 3) +
  geom_vline(xintercept = 2, colour = "red", linetype = "dashed") +
  scale_x_continuous(breaks = 1:3) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Percent Variance Explained",
    subtitle = "K = 2 explains ~80% of the variance",
    x = "Eigenvector # (Principal Component)",
    y = "% Variance Explained"
  ) +
  theme_classic()
```

```{r}
#create a matrix called V including only eigenvectors 1->k
V <- eigenvectors[, 1:2]
```

```{r}
# dot product - multiply Z with V. these are the principal components. call it PC_M
# a*b = (a1b1)+(a2b2)+...+(anbn)
PC_M <- Z %*% V
```

```{r}
# double check PC_M has same time period as Z

PC_M_df <- as_tibble(PC_M) |> 
  mutate(date = returns_wide$date) |>
  relocate(date) # Moves date to the first column

colnames(PC_M_df) <- c("date", "PC1", "PC2")

head(PC_M_df)
```

```{r}
# run OLS on every asset in portfolio. to do so use a loop.
# follows the structure: model <- lm(Asset1 ~ PC1 + PC2 + PC3, data = df) 
model_data <- returns_wide |> inner_join(PC_M_df, by = "date")

model_goog <- lm(GOOG ~ PC1 + PC2, data = model_data)
model_bg   <- lm(BG ~ PC1 + PC2, data = model_data)
model_rtx  <- lm(RTX ~ PC1 + PC2, data = model_data)

summary(model_goog)
```

```{r}
#install.packages("stargazer")
library(stargazer)

stargazer(model_goog, model_bg, model_rtx, type = "text", title = "PCA Beta Output")
```

# Understanding the Betas:
PC1:overall market trend, pc2 constant
- all 3 move with the market. 
- BG, RTX more sensitive to overall market than GOOG --> likely a size story. 

PC2: Industry trends, pc1 constant 
- when pc2 is positive, tech outperforms non-tech. 
- rtx is closer to tech than bg so has a weaker negative beta.

Constants: pc, pc2 constant
- goog outperforms, but all 3 expected to increase

R squared:
- high explanatory power for goog and bg. 
- low explanatory power for rtx. 
- likely bc government contracts, wars, media, etc. cause rapid shocks that affect defense in ways other sectors don't experience.


# comparison: Investment Applications Theory Applied 

```{r}
#CAPM

```

```{r}
# fara and french 3 factor

```

```{r}
# fara and french 5 factor

```


# comparison: derived from conversation with prof. cheah (FMR-A):
```{r}
#original factor model w/ 3 stocks but add T-Bills
```
 







